---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# knnwtsim

<!-- badges: start -->
[![R-CMD-check](https://github.com/mtrupiano1/knnwtsim/workflows/R-CMD-check/badge.svg)](https://github.com/mtrupiano1/knnwtsim/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/knnwtsim)](https://CRAN.R-project.org/package=knnwtsim)
<!-- badges: end -->

The goal of `knnwtsim` is to provide a package to share and implement a forecasting methodology using k nearest neighbors (KNN) primarily for situations where the response series of interest can be predicted by
a combination of its' own recent realizations, its own periodic patterns, and by the values of
one or multiple exogenous predictors. 

The functions of this package are focused primarily into two components. The first being calculation of a 
similarity measure which takes into account all three factors listed above (`S_w`), and can weight 
the degree to which each component contributed to the overall similarity. The second being the 
usage of this measure to identify neighbors and perform KNN regression. For a more formal discussion of the 
methodology used and the key user facing functions in `knnwtsim` please see Trupiano (2021) <arXiv:2112.06266v1>. 

## Formulation 

### Weighted Similarity Measure `S_w`

The Similarity measure I have formulated for incorporating the three factors listed above is: 

`S_w = alpha*S_t + beta*S_p + gamma*S_x`

Where `S_t` is the metric used to calculate the similarity matrix generated by `StMatrixCalc()` to measure similarity in terms of pure recency between all observations. `alpha` is the weight between 0-1 assigned to this matrix in the calculation of `S_w`.

`S_p` is the metric used to calculate the similarity matrix generated `SpMatrixCalc()` to measure similarity in terms of where each observation falls along a periodic cycle relative to all others. `beta` is the weight between 0-1 assigned to this matrix in the calculation of `S_w`.

`S_x` is the metric used to calculate the similarity matrix generated by `SxMatrixCalc()` to measure similarity between all observations in terms of the values of one or more exogenous predictors associated with a given observation. `gamma` is the weight between 0-1 assigned to this matrix in the calculation of `S_w`.

The function `SwMatrixCalc()` calls each of the previous similarity matrix functions to generate the final matrix to use in `knn.forecast()`.

The weights `alpha`, `beta`, and `gamma` are recommended to be set so that they sum to 1. In this case each element of `S_w` should also be between 0-1, with the diagonal elements being equal to 1. 

#### Component Similarity Details 

Initially `S_t`, `S_p`, and `S_x` are calculated as dissimilarities, `D_t`,`D_p`, and `D_x`. They are then transformed to similarities by the formula `1/(D+1)`, where `D` is any dissimilarity measure. This also ensures each element of the similarity matrices fall in the range `(0,1]`, with 1 representing the greatest similarity, and the values approaching 0 the least.

When `StMatrixCalc()` generates the intermediate dissimilarity matrix using `D_t`, it calls the function `TempAbsDissimilarity()` for each pairwise combination of the input vector. Which provided the time orders of two points `y_i` and `y_j`, where the time orders are `i` and `j`.Then `TempAbsDissimilarity(i,j)` will return the absolute difference between the two, i.e `abs(i - j)`.

When `SpMatrixCalc()` generates the intermediate dissimilarity matrix using `D_p`, it calls the function `SeasonalAbsDissimilarity()` for each pairwise combination of the input vector. This function takes the values corresponding to the periods of two points in an overall periodic/seasonal cycle, and the total number of periods in a cycle. Where we have two points `y_i` and `y_j` with `p_i` and `p_j` representing the corresponding period of those points, and a total of `p_max` periods in a full periodic cycle. The value returned by `SeasonalAbsDissimilarity(p_i,p_j,p_max)` will be `min(DirectDis,AroundDis)` where `DirectDis <- abs(p_i - p_j)` and `AroundDis <- abs(min(p_i,p_j) - 1) + abs(p_max - max(p_i,p_j)) + 1`. 

This formulation is based on the idea that generally the periods at the very end and very beginning of a cycle should be fairly similar. To clarify through an example, in a monthly cycle where 1 represents January and 12 represents December, these two months are generally more similar to each-other than either are to July at period 7. Continuing with this monthly example, if we have three points which occur in January, March, and November: `p_i = 1`, `p_j = 3`, `p_k = 11` with `p_max = 12`. Then `SeasonalAbsDissimilarity(p_i,p_j,p_max) = 2`, `SeasonalAbsDissimilarity(p_i,p_k,p_max) = 2`, and `SeasonalAbsDissimilarity(p_j,p_k,p_max) = 4`.

Finally, `SxMatrixCalc()` uses the `stats::dist()` function to produce the intermediate dissimilarity matrix using `D_x`, for an input matrix or vector using the method indicated by the `XdistMetric` argument of `SxMatrixCalc()`. Naturally, this limits `D_x` to the methods available in `stats::dist()`.


### KNN Forecasting 

K Nearest Neighbors forecasting is implemented in this package through the function `knn.forecast()`. Using a provided similarity matrix, which is not required to be calculated using `S_w` specifically, the function will perform K Nearest Neighbors regression on each point in a specified index `f.index.in`, returning the mean of the identified, `k.in`, neighbors in the response series `y.in`.  

Mathematically the estimate for a given point `y_t` is formulated as the mean of the previous points in the series `y_i` identified to be in the neighborhood of `y_t`, `K(y_t)`. 

In the neighborhood `K(y_t)` will be the `k` observations of `y_i` with the highest similarity to `y_t` of all eligible members of the time series, meaning `i < t`. Currently in `knn.forecast()` this eligibility constraint is enforced by only considering the rows of the similarity matrix at indices which are not present in `f.index.in` when selecting neighbors while performing KNN regression on each point in `f.index.in`, thus preventing the corresponding points in `y.in` at those indices from being selected as neighbors. Furthermore, rows and columns of `Sim.Mat.in` at indices greater than the maximum value of `f.index.in` will be removed. 

## Installation

You can install the most recent stable version from CRAN with 

``` r
install.packages("knnwtsim")
```

You can install the development version of knnwtsim from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mtrupiano1/knnwtsim")

```

## Example with Known Similarity Matrix Weights and `k`

This is a basic example which shows a full forecasting workflow if the weights 
to use in the generation of `S_w`, and the hyperparameter `k` are known:

```{r generate Sw example}
library(knnwtsim)
## basic example code
data("simulation_master_list")
series.index <- 15
ex.series <- simulation_master_list[[series.index]]$series.lin.coef.chng.x

#Weights pre tuned by random search. In alpha, beta, gamma order   
pre.tuned.wts <- c(0.2148058, 0.2899638, 0.4952303)
pre.tuned.k <- 5

df <- data.frame(ex.series)
#Generate vector of time orders
df$t <- c(1:nrow(df))

#Generate vector of periods
nperiods <- simulation_master_list[[series.index]]$seasonal.periods
df$p <- rep(1:nperiods,length.out=nrow(df))

#Pull corresponding exogenous predictor(s)
X <- as.matrix(simulation_master_list[[series.index]]$x.chng)
XdistMetric <- 'euclidean'

#Number of points to set aside for validation
val.len <- ifelse(nperiods==12,nperiods,nperiods*2)

#Calculate the weighted similarity matrix Sw 
Sw.ex <- SwMatrixCalc(#For the recency similarity St
                      t.in = df$t
                      #For the periodic similarity Sp
                      ,p.in=df$p, nPeriods.in = nperiods
                      #For the exogenous similarity Sx
                      ,X.in=X, XdistMetric.in = XdistMetric
                      #Weights to be applied to each similarity
                      ,weights=pre.tuned.wts)

#View the top corner of the weighted similarity matrix Sw 
cat('\n Dimensions and Slice of S_w \n')
print(dim(Sw.ex))
print(Sw.ex[1:5,1:5])


```

```{r knn-forecast-example}

#Index we want to forecast 
val.index <- c((length(ex.series) - val.len + 1):length(ex.series))

#Generate the forecast 
knn.frcst <- knn.forecast(Sim.Mat.in = Sw.ex
                          ,f.index.in = val.index
                          ,k.in=pre.tuned.k
                          ,y.in=ex.series)

ts.plot(ex.series,ylab="Simulated Time Series Value")
lines(x=val.index,y=knn.frcst,col='red',lty=2)
legend('bottomleft',legend=c('actuals','KNN Forecast'),col=c('black','red'),lty=c(1,2))


```


## Example with Tuning of Similarity Matrix Weights 

In most cases you will likely want to tune the hyperparameters used in the construction
of `S_w`, and the number of nearest neighbors, `k`, to consider for any given point. There 
are many approaches that can be taken to accomplish this tuning, and many users
may choose to implement their preferred approach. However, for those who want something pre-built
I have included a simple tuning function with the package called `knn.forecast.randomsearch.tuning()`.
This function creates a randomly generated `Grid` of potential sets of hyperparameters `k`,`alpha`,`beta`,`gamma` and generates a forecast of length `test.h` on the last `test.h` points in the series. The best set of parameters 
are selected based on which row of the `Grid` led to the forecast with the lowest MAPE (Mean Absolute Percent Error) over the points in the forecast, and returned as the `weight.opt` and `k.opt` items of the list object returned by the function. Additionally, the 'optimal' similarity matrix to pass to `knn.forecast()` is also returned as `Sw.opt`, as are the testing grid (`Grid`), best test MAPE (`Test.MAPE.opt`), and the full vector of test MAPE values (`MAPE.all`) corresponding to each row in `Grid`.

If you do not want to tune the hyperparameters on the entire series and prefer to leave some points for validation of the tuned result the `val.holdout.len` will remove that many points from the end of the series, and will perform the test forecasts on the `test.h` points at the end of 
the series after the validation observations are removed. 

Using the example series from the previous section, we can reproduce the 'pre-tuned' weights from the 
previous section and generate a forecast using the included tuning function.


```{r tuning-example}

#Calculate component similarity matrices
St.ex <- StMatrixCalc(df$t)
Sp.ex <- SpMatrixCalc(df$p,nPeriods=nperiods)
Sx.ex <- SxMatrixCalc(X)

#Set seed for reproducibility 
set.seed(10)
#Run tuning function 
tuning.ex <- knn.forecast.randomsearch.tuning(grid.len=10**4
                                               ,y.in = ex.series
                                               ,St.in = St.ex 
                                               ,Sp.in = Sp.ex
                                               ,Sx.in = Sx.ex
                                               ,test.h = val.len
                                               ,max.k = NA
                                               ,val.holdout.len=val.len)


cat('\n Tuned Hyperparameters \n')
cat('\n S_w Weights \n')
print(tuning.ex$weight.opt)
cat('\n k \n')
print(tuning.ex$k.opt)

#Pull out tuned S_w and k
k.opt.ex <- tuning.ex$k.opt
Sw.opt.ex <- tuning.ex$Sw.opt


```


```{r tuned-knn-forecast-example}

#Generate the forecast 
knn.frcst.tuned <- knn.forecast(Sim.Mat.in = Sw.opt.ex
                              ,f.index.in = val.index
                              ,k.in=k.opt.ex 
                              ,y.in=ex.series)

ts.plot(ex.series,ylab="Simulated Time Series Value")
lines(x=val.index,y=knn.frcst.tuned,col='purple',lty=3)
legend('bottomleft',legend=c('actuals','KNN Forecast Tuned'),col=c('black','purple'),lty=c(1,3))


```

